How explicit is the Explicit Formula?

### 2016-01-24 paper/paper.tex
fix a number of small (and not) typos

The  term `Explicit Formula' as in our title refers to the genre of formula that expresses {\it arithmetically interesting quantities} in terms of the zeroes of  a related zeta-function or $L$-function.  The first such formula appears towards the end of Bernhard Riemann's short and great paper  ``\"{U}ber die Anzahl der Primzahlen unter einer
gegebenen Gr\"{o}sse," published in the
Monatsberichte der Berliner Akademie,
November 1859.  Subsequently there have been many versions of Explicit Formulas, one such notable variant  proved by Hans von  Mangoldt, in his 1895 article  "Zu Riemann's Abhandlung ``\"{U}ber die Anzahl der Primzahlen unter einer gegebenen Gr\"{o}sse"", Journal f\"{u}r die reine und angewandte Mathematik.  We devote this expository  article in honor of Riemann to a discussion of formulas of this type. Our main aim is to formulate some open questions related to the application of the Explicit Formula to the arithmetic of elliptic curves.

  The {\it arithmetically interesting quantity} that appears in most of the `Explicit Formulas' of analytic number theory is often  a partial sum $F(X)\cdot\sum_{p<X}G(p)$ of locally defined quantities $G(p)$ attached to prime numbers $p$, summed to up to some cutoff value, $p<X$  (and normalized for convenience by an elementary factor $F(X)$). The formula expresses this function of the variable $X$, $F(X)\cdot\sum_{p<X}G(p)$, as a {\it dominant term}, plus an easily controlled error term, plus an interesting third term that might be called the {\it oscillatory term} that, in most cases, is only conjecturally controllable.

@@

  Specifically we will  focus on {\it issues of bias}  following the classical ``Explicit Formula,'' and the work of:
Sarnak, \ Granville,\  Rubenstein,\, Martin-Watkins,\ Fiorilli,\ Bober,\ Conrey-Snaith, and others.
 The example-problem we consider is related to the question---given an elliptic curve over the rational numbers and letting  $p$ range through prime numbers---of how often  $p+1$ is an {\it over-count} or an {\it under-count} for the number of rational points on the curve modulo $p$? The rough answer is 50/50, but there can be a `bias' similar to the classical Chebyshev bias  \cite{R-S}. For such finer statistics one resorts to `Explicit Formulas.'  Here, computation can even outstrip theory in that people have algorithms to make such computations whether or not the holomorphicity of the $L$-functions in question have been proved. Computations have depended on the work of many people.

@@

\section{Brief Introduction}
 One of us (B.M.) having recently taught the classical {\it Explicit Formula} in a standard graduate course in analytic number theory, and having proved that eponymous formula, garnished---as it usually is--- by a number of so-called ``effective constants," $c_1, c_2,$ etc., felt  that, for some applications, this lettered effectiveness left us still too far from  the statistical phenomena behind the formula. To get a closer bead on things, one would do well to work with the numbers behind these   $c_1$'s and $c_2$'s, etc.  It is natural then to see how `actual data'---cutting off the terms of the formula at suitably large values of $X$ given the range of currently feasible computations---compare with the expected results for arbitrarily large $X$.

@@

 We offer no new theoretical results but use this occasion to mention some interesting recent work and conjectures  (of other people)  that might warrant more such computations and that raise a host of questions, both theoretical and computational.   For example, to do some systematic numerical computations related to an elliptic curve $E$ attached to a newform $f_E$  (along the lines of what has already been done in this paper)  it would be very useful to have a much larger data-set  of the arithmetic function  $$n \mapsto r_E(n)$$
 where $r_E(n)$ is the order of vanishing of the $L$-function of the automorphic forms $symm^nf_E$ for odd values of $n$.  Regarding this arithmetic function, aside from having control of the parity of   $r_E(n)$  (e.g., see \cite{DMW}))  hardly anything else is known. Nor do we (at least, the authors of this paper)  yet have enough experience---when $E$ has no complex multiplication---even to formulate a proper conjecture.

@@

{\Large $${\rm{\it Sum\ of\ local\ data\  }}  \ = \  {\rm{\it Global\ data}}\ +\  {\rm{\it  Easy\ error\ term}}\ +\  {\rm{\it  Oscillatory\ term}}.$$ }

@@

  \begin{itemize} \item We require the value $G(p)$ to be determined by only {\it local }considerations at the prime $p$.
The simplest example of such a {\it Sum\ of\ local\ data}  is given by taking $F(X)$ to be the constant $1$, and  $G(p)$ to be  $1$ for all $p$, giving us the classical $\pi(X):=\sum_{p\le X}1$. \item The normalizing factors $F(X)$ will be elementary smooth functions of the cutoff $X$. We sometimes choose this normalizing factor so that the  resulting  $\delta(X)$ has (conjecturally) finite mean.\end{itemize}

We will be concentrating on  {\it sums\ of\ local\ data}  attached to elliptic curves over ${\Q}$,   $$\delta_E(X):=F(X)\sum_{p\le X}g_E(p)$$ where the weighting function $$p \mapsto g_E(p)$$   is a function of $a_E(p)$, the $p$-th Fourier coefficient of the eigenform of weight two parametrizing the elliptic curve.

  We will specifically be interested in  issues of bias. This is what we mean: thanks to the recent resolution of the Sato-Tate Conjecture in this context, one knows that---roughly---half the Fourier coefficients  $a_E(p)$ are positive and half negative.  Indeed, the numbers of positive values and negative values look very close:
\vskip20pt

@@

\vskip20pt
  To study, then, the weighted sums that directly reflect finer statistical issues related to this symmetric distribution, we will be concentrating on weighting functions $p \mapsto g_E(p)$ that have the property that \begin{itemize} \item for all primes $p$, $g_E(p)$ is an {\it odd} function of the value  $a_E(p)$, and \item the {\it sum\ of\ local\ data}  $$\delta_E(X):=\sum_{p\le X}g_E(p)$$ has---or can be convincingly conjectured to have---a finite mean{\footnote{ See Section {\ref{mean}} below}} relative to multiplicative measure $dX/X$.\end{itemize}  In such a context  the mean of $\delta_E(X)$ can be interpreted as a {\it bias}!

 For example, to consider  the problem highlighted in our {\it Preface} (above) form  the `sum of local data'
  $${\frac{\log X}{{\sqrt X}}}\sum_{p\le X}\gamma_E(p)$$  where $\gamma_E(p)=0$ if $p$ is a bad or supersingular prime for $E$ and is otherwise is $+1$ if $E$ has less that $p+1$ rational points over ${\bf F}_p$; and $\gamma_E(p) = -1$ if more.  Then this sum, which will be denoted $\Delta_E(X)$ below, measures exactly the difference between over-count and under-count, as formulated in the {\it Preface}.\vskip10pt
\section{ The three terms on the RHS  of these Explicit Formulas}
  \vskip10pt
\item The first  term on the RHS of our Explicit Formulas, i.e. the term we labeled ${\rm{\it Global\ data}}$, is a constant independent of $X$   determined by the certain `central'  (real)  zeros, or the poles, of the relevant $L$-function or collection of $L$-functions. The   collection of $L$-functions {\it relevant} to the problem highlighted in our {\it Preface} consists of  the $L$ functions attached to {\it all}  the {\it odd} symmetric powers of the newform $f_E$, and the Global data is a function of the multiplicity of the zeroes of these $L$-functions at their central (real) points, where central refers to the functional equation that they enjoy.

Often, and under GRH, these {\it real, central} zeroes of the relevant $L$-functions will have---conjecturally---a clean {\it global arithmetic interpretation} (e.g., via BSD and its  variants) so that's why we call  that term simply  ``{\rm{\it Global\ data}}."  In the problems we will be discussing this ``global data" will be showing up as a certain {\it bias} in the arithmetic statistics of elliptic curves that hearkens back to the early work of Birch and Swinnerton-Dyer, but in the context of the vocabulary we will be using, was  first written down by Peter Sarnak; this is in the spirit of the classical Chebyschev bias, and 'prime races;' an `Explicit Formula' account of this classical theory can be found in {\cite{GM}}.
\item Often the
'easy error term' converges (like $O(1/
log X)$) to a value  (perhaps zero) as $X$ tends to infinity.
\item The {\rm{\it  Oscillatory\ term}} is determined by the (infinitely many) complex (``nontrivial") zeroes. It is usually  an infinite sum, where  (after appropriate conjectures) the summands are of the form  $X^{i\gamma}/f(\gamma)$ where $\gamma$ runs through the imaginary parts of the complex zeroes of the relevant $L$-functions, and $f(y)$ is some natural function.  Numerically, this oscillatory term will indeed oscillate---as we shall amply see---but often one  is tempted to, at the very least, conjecture some control over this wild card.  In actual computations we are surprised by  how small it is.\end{itemize}

\vskip20pt
 \centerline{ \Large{ Part I: Setting up}}
  \vskip30pt
 \section{Bias Questions} Let $E$ be an elliptic curve over ${\Q}$  with no complex multiplication, associated to a newform whose $p$-th Fourier coefficient for $p$ a prime is denoted, as usual, $a_E(p)$. Given the recent work on Sato-Tate, the probability distribution determined by  the normalized values   ${\frac{a_E(p)}{2{\sqrt {p}}}}$  is known to be symmetric about the origin for a large class of such elliptic curves. To repeat our starting question:

@@

 $${\frac{\#\{p < X \ | \ N_E(p) < p+1 \}} {\#\{p > X \ | \ N_E(p) > p+1 \}}}\ =\ {\frac{\#\{p < X \ | \ a_E(p) > 0 \}} {\#\{p > X \ | \ a_E(p) < 0 \}}}$$

@@

 This type of question, of course, bears on Birch's and  Swinnerton-Dyer's initial ``hunch" that the  statistical preponderance of solutions modulo $p$ of an elliptic curve is a predictor of whether or not the elliptic curve has infinitely many rational points.
 \section{ The LHS of our Explicit Formulas}
  To give some ad hoc terms for variant partial sums of {\it Local Arithmetic Data} that measure such preponderances,  let us refer to\begin{itemize}\item   (the slightly doctored version of) the straight difference,
 $${\Delta}_E(X):=  {\frac{\log\ X}{\sqrt X}}\big(\#\{ p < X\ | \ a_E(p) > 0\}\ - \ \#\{ p < X\ | \  a_E(p) < 0\}\big),$$  as  the {\bf raw data,} \item and to
 $${\mathcal D}_E(X):= {\frac{\log\ X}{\sqrt X}}\sum_{p \le X}{\frac{a_E(p)}{\sqrt p}}$$ as the {\bf medium-rare data}, and \item
  $${D}_E(X):= {\frac{1}{\log\ X}}\sum_{p \le X}{\frac{a_E(p)\log p}{ p}}$$  as the {\bf well-done data}.
  \end{itemize}
 \subsection{The statistical distinctions between the three formats}\label{statdist}
   Not to build up too much suspense here, the reason for selecting these three formats for the ``Local data"  and for the specific normalizations chosen (i.e., the factor $ {\frac{\log\ X}{\sqrt X}}$ occurring in the first two, and the factor  ${\frac{1}{\log\ X}}$ in the third)  is that they each are amenable to analysis via ``an" {\it Explicit Formula}
   $$(*)\  \  \  \  {\rm{\it Sum\ of\ local\ data\  }}  \ = \  {\rm{\it Global\ data}}\ + \  {\rm{\it  Easy\ error\ term}}\ + \  {\rm{\it  Oscillatory\ term}}$$
 and such that if (GRH plus)  certain interesting conjectures hold---then  all three  Sums of Local Data, $${\Delta}_E(X), \ {\mathcal D}_E(X), \ {\rm and} \ D_E(X)$$ will have finite {\it means}  (relative to the measure $dX/X$ on ${\bf R}^+$), their `means' being equal to the term  {\rm{\it Global\ data}} in their corresponding Explicit Formula;  and furthermore, what distinguishes these three formats is that conjecturally{\footnote{ as described in a letter of Sarnak; see subsection{\ref{SL}} below.}}---
   \begin{itemize}
   \item the raw data will have {\it infinite} variance,
   \item the medium-rare data will have {\it finite variance}, and
   \item the well-done data will actually achieve its mean as a limiting value.
   \end{itemize}

>>>>

The  term `Explicit Formula' as in our title refers to the genre of formula that expresses {\it arithmetically interesting quantities} in terms of the zeroes of  a related zeta-function or $L$-function.  The first such formula appears towards the end of Bernhard Riemann's short and great paper  ``\"{U}ber die Anzahl der Primzahlen unter einer
gegebenen Gr\"{o}sse," published in the
Monatsberichte der Berliner Akademie,
November 1859.  Subsequently there have been many versions of Explicit Formulas, one such notable variant  proved by Hans von  Mangoldt, in his 1895 article  ``Zu Riemann's Abhandlung `\"{U}ber die Anzahl der Primzahlen unter einer gegebenen Gr\"{o}sse'", Journal f\"{u}r die reine und angewandte Mathematik.  We devote this expository  article in honor of Riemann to a discussion of formulas of this type. Our main aim is to formulate some open questions related to the application of the Explicit Formula to the arithmetic of elliptic curves.

  The {\it arithmetically interesting quantity} that appears in most of the `Explicit Formulas' of analytic number theory is often  a partial sum $F(X)\cdot\sum_{p<X}G(p)$ of locally defined quantities $G(p)$ attached to prime numbers $p$, summed to up to some cutoff value, $p<X$  (and normalized for convenience by an elementary factor $F(X)$). The formula expresses this function $F(X)\cdot\sum_{p<X}G(p)$ of the variable $X$ as a {\it dominant term}, plus an easily controlled error term, plus an interesting third term that might be called the {\it oscillatory term} that, in most cases, is only conjecturally controllable.

@@

  Specifically we will  focus on {\it issues of bias}  following the classical ``Explicit Formula,'' and the work of:
Sarnak, \ Granville,\  Rubenstein,\, Martin-Watkins,\ Fiorilli,\ Bober,\ Conrey-Snaith, and others.
 The example problem we consider is related to the question---given an elliptic curve over the rational numbers and letting  $p$ range through prime numbers---of how often  $p+1$ is an {\it over-count} or an {\it under-count} for the number of rational points on the curve modulo $p$? The rough answer is 50/50, but there can be a `bias' similar to the classical Chebyshev bias  \cite{R-S}. For such finer statistics one resorts to `Explicit Formulas.'  Here, computation can even outstrip theory in that people have algorithms to make such computations whether or not the holomorphicity of the $L$-functions in question have been proved. Computations have depended on the work of many people.

@@

\section{Brief Introduction}
 One of us (B.M.) having recently taught the classical {\it Explicit Formula} in a standard graduate course in analytic number theory, and having proved that eponymous formula, garnished---as it usually is---by a number of so-called ``effective constants," $c_1, c_2,$ etc., felt  that, for some applications, this lettered effectiveness left us still too far from  the statistical phenomena behind the formula. To get a closer bead on things, one would do well to work with the numbers behind these   $c_1$'s and $c_2$'s, etc.  It is natural then to see how `actual data'---cutting off the terms of the formula at suitably large values of $X$ given the range of currently feasible computations---compare with the expected results for arbitrarily large $X$.

@@

 We offer no new theoretical results but use this occasion to mention some interesting recent work and conjectures  (of other people)  that might warrant more such computations and that raise a host of questions, both theoretical and computational.   For example, to do some systematic numerical computations related to an elliptic curve $E$ attached to a newform $f_E$  (along the lines of what has already been done in this paper)  it would be very useful to have a much larger data set  of the arithmetic function  $$n \mapsto r_E(n)$$
 where $r_E(n)$ is the order of vanishing of the $L$-function of the automorphic forms ${\rm symm}^n f_E$ for odd values of $n$.  Regarding this arithmetic function, aside from having control of the parity of   $r_E(n)$  (e.g., see \cite{DMW}))  hardly anything else is known. Nor do we (at least, the authors of this paper)  yet have enough experience---when $E$ has no complex multiplication---even to formulate a proper conjecture.

@@

{\Large $${\text{Sum of local data}}  \ = \  {\text{Global data}}\ +\  {\text{Easy error term}}\ +\  {\text{Oscillatory term}}.$$ }

@@

  \begin{itemize} \item We require the value $G(p)$ to be determined by only {\it local }considerations at the prime~$p$.
The simplest example of such a {\it Sum\ of\ local\ data}  is given by taking $F(X)$ to be the constant $1$, and  $G(p)$ to be  $1$ for all $p$, giving us the classical prime counting function $\pi(X):=\sum_{p\le X}1$. \item The normalizing factors $F(X)$ will be elementary smooth functions of the cutoff $X$. We sometimes choose this normalizing factor so that the  resulting  $\delta(X)$ has (conjecturally) finite mean.\end{itemize}

We will be concentrating on  {\it sums\ of\ local\ data}  attached to elliptic curves over ${\Q}$,   $$\delta_E(X):=F(X)\sum_{p\le X}g_E(p)$$ where the weighting function $$p \mapsto g_E(p)$$   is a function of $a_E(p)$, the $p$th Fourier coefficient of the eigenform of weight two parametrizing the elliptic curve.

  We will specifically be interested in  issues of bias. This is what we mean: thanks to the recent resolution \cite{} of the Sato-Tate Conjecture in this context, one knows that---roughly---half the Fourier coefficients  $a_E(p)$ are positive and half negative (for non-CM curves).  Indeed, the numbers of positive values and negative values look very close to each other:
\vskip20pt

@@

\vskip20pt
  To study, then, the weighted sums that directly reflect finer statistical issues related to this symmetric distribution, we will be concentrating on weighting functions $p \mapsto g_E(p)$ that have the property that \begin{itemize} \item for all primes $p$, $g_E(p)$ is an {\it odd} function of the value  $a_E(p)$, and \item the {\it sum\ of\ local\ data}  $$\delta_E(X):=\sum_{p\le X}g_E(p)$$ has---or can be convincingly conjectured to have---a finite mean{\footnote{ See Section~\ref{mean} below.}} relative to multiplicative measure $dX/X$.\end{itemize}  In such a context  the mean of $\delta_E(X)$ can be interpreted as a {\it bias}!

 For example, to consider  the problem highlighted in our {\it Preface} (above) form  the `sum of local data'
  $${\frac{\log X}{{\sqrt X}}}\sum_{p\le X}\gamma_E(p)$$  where $\gamma_E(p)=0$ if $p$ is a bad or supersingular prime for $E$ and is otherwise is $+1$ if $E$ has less that $p+1$ rational points over ${\bf F}_p$; and $\gamma_E(p) = -1$ if more.  Then this sum, which will be denoted $\Delta_E(X)$ below, measures exactly the difference between over-count and under-count, as formulated in the {\it Preface}.\vskip10pt
\section{ The three terms on the right hand side (RHS)  of these Explicit Formulas}
  \vskip10pt
\item The first  term on the RHS of our Explicit Formulas, i.e. the term we labeled ${\rm{\it Global\ data}}$, is a constant independent of $X$   determined by the certain `central'  (real)  zeros, or the poles, of the relevant $L$-function or collection of $L$-functions. The   collection of $L$-functions {\it relevant} to the problem highlighted in our {\it Preface} consists of  the $L$ functions attached to {\it all}  the {\it odd} symmetric powers of the newform $f_E$, and the Global data is a function of the multiplicity of the zeroes of these $L$-functions at their central (real) points, where central refers to the functional equation that they enjoy.

Often, and under GRH, these {\it real, central} zeroes of the relevant $L$-functions will have---conjecturally---a clean {\it global arithmetic interpretation} (e.g., via BSD and its  variants) so that's why we call  that term simply  ``{\rm{\it Global\ data}}."  In the problems we will be discussing this ``global data" will be showing up as a certain {\it bias} in the arithmetic statistics of elliptic curves that hearkens back to the early work of Birch and Swinnerton-Dyer \cite{}, but in the context of the vocabulary we will be using, was  first written down by Peter Sarnak; this is in the spirit of the classical Chebyschev bias, and 'prime races;' an `Explicit Formula' account of this classical theory can be found in {\cite{GM}}.
\item Often the
`easy error term' converges (like $O(1/
\log{X})$) to a value  (perhaps zero) as $X$ tends to infinity.
\item The {\rm{\it  Oscillatory\ term}} is determined by the (infinitely many) complex (``nontrivial") zeroes. It is usually  an infinite sum, where  (after appropriate conjectures) the summands are of the form  $X^{i\gamma}/f(\gamma)$ where $\gamma$ runs through the imaginary parts of the complex zeroes of the relevant $L$-functions, and $f(y)$ is some natural function.  Numerically, this oscillatory term will indeed oscillate---as we shall amply see---but often one  is tempted to, at the very least, conjecture some control over this wild card.  In actual computations we are surprised by  how small it is.\end{itemize}

\vskip20pt
 \centerline{ \Large{ Part I: Setting up}}
  \vskip30pt
 \section{Bias Questions} Let $E$ be an elliptic curve over ${\Q}$  with no complex multiplication, associated to a newform whose $p$th Fourier coefficient for $p$ a prime is denoted, as usual, $a_E(p)$. Given the recent work \cite{} on Sato-Tate, the probability distribution determined by  the normalized values   ${\frac{a_E(p)}{2{\sqrt {p}}}}$  is known to be symmetric about the origin for a large class of such elliptic curves. To repeat our starting question:

@@

 $${\frac{\#\{p < X \ | \ N_E(p) < p+1 \}} {\#\{p < X \ | \ N_E(p) > p+1 \}}}\ =\ {\frac{\#\{p < X \ | \ a_E(p) > 0 \}} {\#\{p < X \ | \ a_E(p) < 0 \}}}$$

@@

 This type of question, of course, bears on Birch's and  Swinnerton-Dyer's initial ``hunch" that the  statistical preponderance of solutions modulo $p$ of an elliptic curve is a predictor of whether or not the elliptic curve has infinitely many rational points.
 \section{ The left hand side (LHS) of our Explicit Formulas}
  To give some ad hoc terms for variant partial sums of {\it Local Arithmetic Data} that measure such preponderances,  let us refer to\begin{itemize}\item   (the slightly doctored version of) the straight difference,
 $${\Delta}_E(X):=  {\frac{\log\ X}{\sqrt X}}\big(\#\{ p < X\ | \ a_E(p) > 0\}\ - \ \#\{ p < X\ | \  a_E(p) < 0\}\big),$$  as  the {\bf raw data,} \item and to
 $${\mathcal D}_E(X):= {\frac{\log\ X}{\sqrt X}}\sum_{p \le X}{\frac{a_E(p)}{\sqrt p}}$$ as the {\bf medium-rare data}, and \item
  $${D}_E(X):= {\frac{1}{\log\ X}}\sum_{p \le X}{\frac{a_E(p)\log p}{ p}}$$  as the {\bf well-done data}.
  \end{itemize}
 \subsection{The statistical distinctions between the three formats}\label{statdist}
   Not to build up too much suspense here, the reason for selecting these three formats for the ``Local data"  and for the specific normalizations chosen (i.e., the factor $ {\frac{\log\ X}{\sqrt X}}$ occurring in the first two, and the factor  ${\frac{1}{\log\ X}}$ in the third)  is that they each are amenable to analysis via ``an" {\it Explicit Formula}
   $$(*)\  \  \  \  {\rm{\it Sum\ of\ local\ data\  }}  \ = \  {\rm{\it Global\ data}}\ + \  {\rm{\it  Easy\ error\ term}}\ + \  {\rm{\it  Oscillatory\ term}}$$
 and such that if (GRH plus)  certain interesting conjectures hold---then  all three  Sums of Local Data, $${\Delta}_E(X), \ {\mathcal D}_E(X), \ {\rm and} \ D_E(X)$$ will have finite {\it means}  (relative to the measure $dX/X$ on ${\bf R}^+$), their `means' being equal to the term  {\rm{\it Global\ data}} in their corresponding Explicit Formula;  and furthermore, what distinguishes these three formats is that conjecturally{\footnote{ as described in a letter of Sarnak; see subsection{\ref{SL}} below.}}---
   \begin{itemize}
   \item the raw data will have {\it infinite} variance,
   \item the medium-rare data will have {\it finite variance}, and
   \item the well-done data will actually achieve its mean as a limiting value.
   \end{itemize}
